Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-03-15 analysis/paper/paper.Rmd
note to self
reshaping the discussion

\
Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].

1.  Consequences for demographers

    1.  Very few demographic studies have the data required to fit models that include delayed effects.

    2.  This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
        We found evidence for delayed effects of drought which are sometimes better predictors of vital rates than more recent conditions.
        Delayed effects of climate on plant vital rates may be the rule, rather than the exception, and distributed lag non-linear models offer an unbiased way of identifying these delayed effects (but see @pierreCriticalWindowsMethod2020 for an alternative method).
        <!--# maybe the previous sentence goes in methods? -->

2.  Consequences for biology

    1.  What look like small changes in vital rates add up over time, especially for slow-growing, long-lived organisms [@gagnonGrowthUnderstoryHerb2011].

    2.  Average flowering probability of plants in continuous forest is equivalent to flowering once every \_\_\_ years compared to once every \_\_\_ years in 1-ha fragments (assuming plant remains the same size and SPEI is average).
        <!--# from model intercepts -->

        1.  Previous work has shown flowering probability to be an important vital rate. Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.

    3.  Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability.

        1.  The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018].

        2.  Alternatively, the delayed effect of drought could be due to indirect effects mediated by canopy openness.
            In Manaus, canopy is as green or greener in dry season compared to wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought has the potential to increase solar radiation if it causes reductions in canopy cover.
            Tree mortality (CITATIONS) and litterfall (CITATIONS) in the Amazon increase following drought, which could increase solar radiation to the understory resulting in greater energy capture by *H. acuminata*.

        3.  More work to be done on flowering responses to drought x light x season interactions.

    4.  Address how all but the smallest sizes are shrinking on average over the period of this study (Figure \@ref(fig:growth-size)).
        Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.

3.  Buffering

@@

Future directions/Caveats:

@@

# Conclusion

>>>>

\
1. Understanding how landscape structure and environmental conditions act to influence population dynamics is fundamental to many of the emerging conceptual frameworks for studying and conserving fragmented landscapes [@didhamRethinkingConceptualFoundations2012; @driscollConceptualDomainMatrix2013].

2. Further evidence delayed effects may be the rule (emphasis on demographic modeling)

      * This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021]. 
      * We found evidence for delayed effects of drought which are sometimes better predictors of vital rates than more recent conditions. 
      * Delayed effects of climate on plant vital rates may be the rule, rather than the exception. 
      * distributed lag non-linear models offer an unbiased way of identifying these delayed effects (but see @pierreCriticalWindowsMethod2020 for an alternative method). <!--# maybe the previous sentence goes in methods? -->

3. What makes our results important is that we went the next step: we not only found that there were delayed effects, but that these effects different by habitat and life-history stage.  [this lays out what these patterns are]

- Growth v Survivorship v Flowering

- What look like small changes in vital rates add up over time, especially for slow-growing, long-lived organisms [@gagnonGrowthUnderstoryHerb2011].

 - Average flowering probability of plants in continuous forest is equivalent to flowering once every \_\_\_ years compared to once every \_\_\_ years in 1-ha fragments (assuming plant remains the same size and SPEI is average). <!--# from model intercepts -->

- Previous work has shown flowering probability to be an important vital rate. Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.

- Two dry seasons ago (lag of 15--20 months) is important for all three vital rates, but most clearly for flowering where drought in this period increases flowering probability.

- The increase in flowering rate of plants following a drought could be a bet-hedging strategy in response to stress [@nihadRelationshipStressFlowering2018].

4. Mechanisms [this lays out the biological, fragmentation-related mechanisms underpinning these patterns]

- I gather that in a good year, plants can store carbon and nutrients in rhizomes and then mobilize as needed and the opposite is true in a bad year. Flowering is a big investment so perhaps flowering in one year means that you don’t flower the following one. If that is the case, it would make sense to include growth in the previous year as a covariate in all analyses rather than do this indirectly via a potential effect of past SPEI on past growth. It will also be easier to explain as you write the paper. **Emilio will look into physiological responses to light and H20 in the horticultural and ecological literature, eg. Westerband papers, esp if focusing on herbs** Basically anything on how more or less precip influences vital rates or how plants respond to them, esp if linked to changes in canopy openness, litterfall, flowering in ENSO years, Bruna transplant experiments, Heliconia germination in secondary forest, important to emphasize: results are often counter-intuitive.

5. Ecological Consequences of these patterns: an eye towards a drier future

- Alternatively, the delayed effect of drought could be due to indirect effects mediated by canopy openness. In Manaus, canopy is as green or greener in dry season compared to wet season [@guanPhotosyntheticSeasonalityGlobal2015], so dry season drought has the potential to increase solar radiation if it causes reductions in canopy cover. Tree mortality (CITATIONS) and litterfall (CITATIONS) in the Amazon increase following drought, which could increase solar radiation to the understory resulting in greater energy capture by *H. acuminata*.

-  More work to be done on flowering responses to drought x light x season interactions.

- Address how all but the smallest sizes are shrinking on average over the period of this study (Figure \@ref(fig:growth-size)). Cite: @brunaExperimentalAssessmentHeliconia2002; Begg 1980; Wright 1996; Rundel et al. 1998.

6.  Buffering [where does this fit best...as a mechanism or something we need to pin down with more work?]

@@

@@

# Conclusions & Future Directions

-  Very few demographic studies have the data required to fit models that include delayed effects.

-  Looop back to: need whole-life-cycle approach to ID these effects, why fragmentologists don't do this is killing Emilio.
